\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-12-21 misc/JSS article/final/article.Rnw
Update package website url to fastverse.org/collapse.

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://fastverse.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://fastverse.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, including vectorized and verbose hash-joins and fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications of the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline

@@

The package comes with a built-in structured \href{https://fastverse.github.io/collapse/reference/index.html}{documentation} facilitating its use. This includes a central \href{https://fastverse.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and a set of supplementary topic pages which briefly summarize related functionality. The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:

@@

While this article cannot fully present \pkg{collapse}, the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

@@

The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R}/\pkg{collapse} means that these operations are implemented using compiled \proglang{C/C++} code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data. Their basic syntax is

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - sampling weights (only some functions), and \code{TRA} - transformation of \code{x}. The following examples with \fct{fmean} demonstrate their basic usage on the familiar \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris}{\code{iris}} dataset recording 50 measurements of 4 variables for 3 species of iris flowers.\footnote{\fct{num\_vars} returns the numeric variables in a data frame-like object; \fct{cat\_vars} the categorical ones.} All examples support weights (\code{w}), and \fct{fmean} can also be multithreaded across columns (\code{nthreads}).\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions, as detailed in the respective function documentation. The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. % Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()

@@

Having grouping and data transformation functionality directly built into generic statistical functions facilitates and speeds up many common operations. It notably avoids the need to convert atomic objects to data frames for grouped aggregations or transformations. The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://fastverse.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:

@@

\pkg{collapse} also provides a broad set of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc. as optimized analogues to base \proglang{R}/\pkg{tidyverse} functions. These are integrated with the \emph{Fast Statistical and Transformation Functions} to enable vectorized operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow. I illustrate this using the included World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.

@@

\pkg{collapse} also vectorizes advanced statistics. The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} computes weighted quantiles in a theoretically consistent way, see \href{https://fastverse.github.io/collapse/reference/fquantile.html}{fquantile} for details.}

@@

 Below, I create a \class{GRP} object from two columns in the World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}). The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.

@@

The lower-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://fastverse.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}. Users should note that \pkg{collapse} does not provide metaprogramming capabilities in its non-standard evaluation functions---such as \href{https://rlang.r-lib.org/reference/topic-quosure.html}{quosures} or \href{https://dplyr.tidyverse.org/articles/programming.html}{indirection} familiar to \pkg{tidyverse} users. Instead, it has standard-evaluation equivalents to some of these functions which typically end with a \code{v} for 'variables', such as \fct{collapv}, \fct{fslicev}, \fct{ftransformv}, etc. In other cases, including those typically handled by \fct{fsummarise} or \fct{fmutate}, users need to use the lower-level API for programming, or resort to \fct{substitute} and friends. The main reason for not providing higher-level metaprogramming capabilities is to keep all functions as simple as possible. It also compels users to think deeply about their programs and devise more efficient solutions. \newline

@@

to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former. Any data frame-like object can become an \class{indexed\_frame} and behave as usual for other operations. The technical implementation of these classes is described in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://fastverse.github.io/collapse/reference/indexing.html}{documentation}. Their basic usage is:

@@

Data can be indexed using one or more indexing variables. Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure---every series inside the frame is already an \class{indexes\_series}. A comprehensive set of \href{https://fastverse.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()} ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.). % A basic demonstration with World Bank panel data showcases the flexibility of these classes.

@@

An example follows using the \code{exports} data. \emph{Note} that data can be unsorted for indexation.
<<indexing>>=
exportsi <- exports |> findex_by(c, s, y)
exportsi |> G(0:1) |> head(5)
exportsi |> findex() |> print(2)
@
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
<<indexing_2>>=
vi <- exportsi$v; str(vi, width = 70, strict = "cut")
is_irregular(vi)
vi |> psmat() |> head(3)
fdiff(vi) |> psmat() |> head(3)
@
\fct{psmat}, for panel-series to matrix, generates a matrix or array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
<<indexing_3>>=
settransform(exportsi, v_ld = Dlog(v))
lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
@
Indexed series/frames also support transformations such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}. Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series. The \class{index\_df} attached to these objects can also be used with other general tools such as \fct{collapse::BY} to perform grouped computations using third-party functions: %For example, below I compute a 5-year rolling average ().
%
<<rollmean>>=
BY(vi, findex(vi)$c.s, data.table::frollmean, 5) |> head(10)
@
%
Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among its suite of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}). Thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

@@

\fct{collapse::pivot} provides a powerful new implementation of reshaping for \proglang{R} that addresses these shortcomings. It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information. Its basic syntax is: \newpage
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, FUN = "last", ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://fastverse.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
<<pivots>>=
data <- GGDC10S |>
 fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
 fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
@
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels stored in \code{attr(column, "label")} are saved to an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
<<pivot_longer>>=
head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
  labels = "Sector", how = "longer"))
@
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", how = "wider")}, where \code{dl} is the long data. \newline

@@

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. A small number extremely fast internal aggregation functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}, operate 'on the fly' during reshaping. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but requires a deep copy of the columns aggregated which is avoided by the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
<<pivot_agg>>=
head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"),
  how = "recast", names = list(from = "Variable", to = "Sectorcode"),
  labels = list(from = "Label", to = "Sector"), FUN = "mean"))
@
The \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}. Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects, such as a model predictions and performance statistics, in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

>>>>

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://fastverse.org/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://fastverse.org/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, including vectorized and verbose hash-joins and fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://fastverse.org/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications of the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline

@@

The package comes with a built-in structured \href{https://fastverse.org/collapse/reference/index.html}{documentation} facilitating its use. This includes a central \href{https://fastverse.org/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and a set of supplementary topic pages which briefly summarize related functionality. The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:

@@

While this article cannot fully present \pkg{collapse}, the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.org/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.org/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

@@

The \href{https://fastverse.org/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R}/\pkg{collapse} means that these operations are implemented using compiled \proglang{C/C++} code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data. Their basic syntax is

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - sampling weights (only some functions), and \code{TRA} - transformation of \code{x}. The following examples with \fct{fmean} demonstrate their basic usage on the familiar \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris}{\code{iris}} dataset recording 50 measurements of 4 variables for 3 species of iris flowers.\footnote{\fct{num\_vars} returns the numeric variables in a data frame-like object; \fct{cat\_vars} the categorical ones.} All examples support weights (\code{w}), and \fct{fmean} can also be multithreaded across columns (\code{nthreads}).\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions, as detailed in the respective function documentation. The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://fastverse.org/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. % Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()

@@

Having grouping and data transformation functionality directly built into generic statistical functions facilitates and speeds up many common operations. It notably avoids the need to convert atomic objects to data frames for grouped aggregations or transformations. The \href{https://fastverse.org/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://fastverse.org/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:

@@

\pkg{collapse} also provides a broad set of \href{https://fastverse.org/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc. as optimized analogues to base \proglang{R}/\pkg{tidyverse} functions. These are integrated with the \emph{Fast Statistical and Transformation Functions} to enable vectorized operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow. I illustrate this using the included World Development Dataset (\href{https://fastverse.org/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.

@@

\pkg{collapse} also vectorizes advanced statistics. The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} computes weighted quantiles in a theoretically consistent way, see \href{https://fastverse.org/collapse/reference/fquantile.html}{fquantile} for details.}

@@

 Below, I create a \class{GRP} object from two columns in the World Development Dataset (\href{https://fastverse.org/collapse/reference/wlddev.html}{\code{wlddev}}). The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.

@@

The lower-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://fastverse.org/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}. Users should note that \pkg{collapse} does not provide metaprogramming capabilities in its non-standard evaluation functions---such as \href{https://rlang.r-lib.org/reference/topic-quosure.html}{quosures} or \href{https://dplyr.tidyverse.org/articles/programming.html}{indirection} familiar to \pkg{tidyverse} users. Instead, it has standard-evaluation equivalents to some of these functions which typically end with a \code{v} for 'variables', such as \fct{collapv}, \fct{fslicev}, \fct{ftransformv}, etc. In other cases, including those typically handled by \fct{fsummarise} or \fct{fmutate}, users need to use the lower-level API for programming, or resort to \fct{substitute} and friends. The main reason for not providing higher-level metaprogramming capabilities is to keep all functions as simple as possible. It also compels users to think deeply about their programs and devise more efficient solutions. \newline

@@

to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former. Any data frame-like object can become an \class{indexed\_frame} and behave as usual for other operations. The technical implementation of these classes is described in the \href{https://fastverse.org/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://fastverse.org/collapse/reference/indexing.html}{documentation}. Their basic usage is:

@@

Data can be indexed using one or more indexing variables. Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure---every series inside the frame is already an \class{indexes\_series}. A comprehensive set of \href{https://fastverse.org/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()} ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.). % A basic demonstration with World Bank panel data showcases the flexibility of these classes.

@@

An example follows using the \code{exports} data. \emph{Note} that data can be unsorted for indexation.
<<indexing>>=
exportsi <- exports |> findex_by(c, s, y)
exportsi |> G(0:1) |> head(5)
exportsi |> findex() |> print(2)
@
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
<<indexing_2>>=
vi <- exportsi$v; str(vi, width = 70, strict = "cut")
is_irregular(vi)
vi |> psmat() |> head(3)
fdiff(vi) |> psmat() |> head(3)
@
\fct{psmat}, for panel-series to matrix, generates a matrix or array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
<<indexing_3>>=
settransform(exportsi, v_ld = Dlog(v))
lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
@
Indexed series/frames also support transformations such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}. Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series. The \class{index\_df} attached to these objects can also be used with other general tools such as \fct{collapse::BY} to perform grouped computations using third-party functions: %For example, below I compute a 5-year rolling average ().
%
<<rollmean>>=
BY(vi, findex(vi)$c.s, data.table::frollmean, 5) |> head(10)
@
%
Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among its suite of \href{https://fastverse.org/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.org/collapse/reference/join.html}{joins} and \href{https://fastverse.org/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}). Thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

@@

\fct{collapse::pivot} provides a powerful new implementation of reshaping for \proglang{R} that addresses these shortcomings. It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information. Its basic syntax is: \newpage
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, FUN = "last", ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://fastverse.org/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
<<pivots>>=
data <- GGDC10S |>
 fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
 fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
@
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels stored in \code{attr(column, "label")} are saved to an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
<<pivot_longer>>=
head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
  labels = "Sector", how = "longer"))
@
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", how = "wider")}, where \code{dl} is the long data. \newline

@@

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. A small number extremely fast internal aggregation functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}, operate 'on the fly' during reshaping. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but requires a deep copy of the columns aggregated which is avoided by the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
<<pivot_agg>>=
head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"),
  how = "recast", names = list(from = "Variable", to = "Sectorcode"),
  labels = list(from = "Label", to = "Sector"), FUN = "mean"))
@
The \href{https://fastverse.org/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}. Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects, such as a model predictions and performance statistics, in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://fastverse.org/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline
